---
title: "Exploratory Data Analysis"
author: "Group 25 - US social determinants of health by county"
date: "11/20/2021"
output:
  html_document:
    toc: yes
    toc_depth: '4'
---

## Setup
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r load libraries, message=FALSE, warning=FALSE}
library(tidyverse)
library(plotly)
library(broom)

```

Our GitHub Repo: 
**https://github.com/UBC-MDS/DSCI_522_US_social_determinants_of_health_by_county**


## Load Data

```{r load datasets for this assignment, message=FALSE}
covid_data <- read.csv("data/US_counties_COVID19_health_weather_data.csv")
```

## Select features

```{r select features for analysis}
interesting_features <- c(
  "date", "county", "cases", "state", 
  "total_population", "num_deaths", "percent_smokers", 
  "percent_vaccinated", "income_ratio", 
  "population_density_per_sqmi", "percent_fair_or_poor_health",
  "percent_unemployed_CHR", "violent_crime_rate", 
  "chlamydia_rate", "teen_birth_rate"
)

covid_data <- covid_data %>% 
  select(all_of(interesting_features)) %>%
  mutate(date = as.Date(date)) # change date from character to "Date" class


# check the descriptive stats of the data frame 
summary(covid_data)
```

## Exploratory Data Analysis (EDA)

### Table of COVID-19 prevalence for every state

```{r}
library(lemon)
knit_print.data.frame <- lemon_print
```

```{r table of covid prevalence for each state, message=FALSE, render=lemon_print}
covid_prevalence_table_state <- covid_data %>%

# The following lines are for calculating daily growth rate
  group_by(state, date) %>%
  summarize(cases = sum(cases),
  population = mean(total_population, na.rm=TRUE)) %>%
  mutate(cases_growth_rate = (cases - lag(cases) / lag(cases))) %>%

  # The following lines are for group_by values for each state
  group_by(state) %>%
  summarize(total_cases = max(cases),
  total_cases_per_capita = total_cases / mean(population, na.rm=TRUE),
  mean_cases_growth_rate = mean(cases_growth_rate, na.rm=TRUE)) %>%
  arrange(desc(total_cases))

head(data.frame(covid_prevalence_table_state))
tail(data.frame(covid_prevalence_table_state))

```
There are `NA`s in the table because of the missing values for that county/state in the original dataset.

### Table of COVID-19 prevalence for every county

```{r table of covid prevalence for each county, message=FALSE, render=lemon_print}
covid_prevalence_table_county <- covid_data %>%
# The following lines are for calculating daily growth rate
  group_by(county, date) %>%
  summarize(cases = sum(cases),
  population = mean(total_population, na.rm=TRUE)) %>%
  mutate(cases_growth_rate = (cases - lag(cases) / lag(cases))) %>%

# The following lines are for group_by values for each state
  group_by(county) %>%
  summarize(total_cases = max(cases),
  total_cases_per_capita = total_cases / mean(population, na.rm=TRUE),
  mean_cases_growth_rate = mean(cases_growth_rate, na.rm=TRUE)) %>%
  arrange(desc(total_cases)) 

head(data.frame(covid_prevalence_table_county))
tail(data.frame(covid_prevalence_table_county))

```

### Visualization 1 - distributions of numeric features

```{r create a dataframe for visualization}
covid_data_group_by_sate <- covid_data %>%
  group_by(state) %>%
  summarize(
            num_deaths = max(num_deaths), 
            percent_smokers = mean(percent_smokers, na.rm=TRUE), 
            percent_vaccinated = max(percent_vaccinated), 
            income_ratio = mean(income_ratio, na.rm=TRUE), 
            population_density_per_sqmi = mean(population_density_per_sqmi, 
                                               na.rm=TRUE), 
            percent_fair_or_poor_health = mean(percent_fair_or_poor_health, 
                                               na.rm=TRUE), 
            percent_unemployed_CHR = mean(percent_unemployed_CHR, na.rm=TRUE), 
            violent_crime_rate = mean(violent_crime_rate, na.rm=TRUE), 
            chlamydia_rate = mean(chlamydia_rate, na.rm=TRUE), 
            teen_birth_rate = mean(teen_birth_rate, na.rm=TRUE)
            ) %>%
  merge(covid_prevalence_table_state, by="state") %>%
  arrange(desc(total_cases)) 
```

```{r density plots for numetric variables, warning=FALSE}


par(mfrow=c(3, 4))

covid_data_group_by_sate_long <- covid_data_group_by_sate %>%
    select_if(is.numeric) %>%
    pivot_longer(everything()) 

covid_data_group_by_sate_long %>%
  ggplot(aes(x=value)) + 
  geom_density(fill='grey') +
  facet_wrap(~name, scales='free') + 
  theme(strip.text = element_text(size=7),
        axis.text.x = element_text(size=5), 
        axis.text.y = element_text(size=5), 
        plot.title = element_text(hjust = 0.5)) +
  labs(title="Density plots of numeric feature",
       x ="", 
       y = "Density") 


```


### Visualization 2 - relationships between total COVID-19 cases per capita of each state and other features

```{r relationships between total COVID-19 cases per capita and other features, warning=FALSE}
par(mfrow=c(3, 4))

covid_data_group_by_sate_long <- covid_data_group_by_sate %>%
    select_if(is.numeric) %>%
    pivot_longer(-total_cases_per_capita) 

case_per_capita_plot <- covid_data_group_by_sate_long %>%
  ggplot(aes(x=value, y=total_cases_per_capita)) + 
  geom_point(alpha = 0.4) +
  facet_wrap(~name, scales='free') +
  theme(strip.text = element_text(size=7),
        axis.text.x = element_text(size=7), 
        axis.text.y = element_text(size=7), 
        plot.title = element_text(hjust = 0.5), 
        panel.spacing = unit(2.5, "lines")) +
  labs(title="Plots of total COVID-19 cases per capita v.s. other features",
       x ="", 
       y = "COVID-19 cases per capita") +
  scale_x_continuous(labels = scales::label_number_si())


fig_case_per_capita_plot <- ggplotly(case_per_capita_plot)
fig_case_per_capita_plot
```



### Visualization 3 - relationships between average COVID-19 cases growth rate for each state and other features

```{r relationships between average COVID-19 cases growth rate and other features, warning=FALSE}
par(mfrow=c(3, 4))

covid_data_group_by_sate_long <- covid_data_group_by_sate %>%
    select_if(is.numeric) %>%
    pivot_longer(-mean_cases_growth_rate) 

covid_growth_rate_plot <- covid_data_group_by_sate_long %>%
  ggplot(aes(x=value, y=mean_cases_growth_rate)) + 
  geom_point(alpha = 0.4) +
  facet_wrap(~name, scales='free') +
  theme(strip.text = element_text(size=7),
        axis.text.x = element_text(size=7), 
        axis.text.y = element_text(size=7), 
        plot.title = element_text(hjust = 0.5), 
        panel.spacing = unit(2.5, "lines")) +
  labs(title="Plots of average COVID-19 growth rate v.s. other features",
       x ="", 
       y = "COVID-19 growth rate") +
  scale_x_continuous(labels = scales::label_number_si()) +
  scale_y_continuous(labels = scales::label_number_si())

fig_covid_growth_rate_plot <- ggplotly(covid_growth_rate_plot)
fig_covid_growth_rate_plot
```





